Indicators for Ecosystem Condition in Norway

### 2023-10-13 functional_plant_indicators_wetland.Rmd

---
output:
  html_document:
    code_folding: show
---

@@

<!-- Load all you dependencies here -->
```{r setup, class.source = 'fold-hide'}
library(sf)
library(plyr)
library(tidyverse)
library(RColorBrewer)
library(gridExtra) 
library(ggridges)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(knitr)
library(kableExtra)
library(betareg)
library(glmmTMB)
knitr::opts_chunk$set(echo = TRUE)
```

@@

Indicators described in this chaper:

@@

## Analyses {#analyses-fpiw}
### Data sets
<!-- Please print the entire workflow with echo=T. Also, simply reading cached data creates loose ends - I cant find out where this data comes from. You may also chose to read tin this data (or a cached version of it) way at the top of the file, so that you can add some in-line R code for the summary stats in the introduction part (see previous comments)-->
ANO data

```{r, class.source = 'fold-hide'}
# Add data from cache
ANO.sp<-readRDS(paste0(here::here(),"/data/cache/ANO.sp.RDS"))
ANO.geo<-readRDS(paste0(here::here(),"/data/cache/ANO.geo.RDS"))

@@

Plant indicators from Tyler et al. (2021)
```{r, class.source = 'fold-hide'}
## Tyler indicator data
## data from cache
ind.dat <-readRDS(paste0(here::here(),"/data/cache/ind.dat.RDS"))
# ind.dat <- read.table("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/Tyler et al_Swedish plant indicators.txt",
#                         sep = '\t', header=T, quote = '')
# head(ind.dat)

@@

Generalized species lists NiN
```{r, class.source = 'fold-hide'}
# data from cache
Eco_State<-readRDS(paste0(here::here(),"/data/cache/Eco_State.RDS"))
# load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/functional plant indicators/reference from NiN/Eco_State.RData")
# str(Eco_State)
```

@@

```{r, include = T, results='hide', warning=F, message=F, class.source = 'fold-hide'}

@@

<!-- Print head of data set with horizontal scrolling -->
```{r ano-sp-tab, class.source = 'fold-hide'}
head(ANO.sp.ind) %>%
  kable("html",caption = "ANO species occurence data set with attached plant trait data.") %>% 
  kable_styling("striped") %>% scroll_box(width = "100%")
```

```{r wetland-ref-data-tab, class.source = 'fold-hide'}
head(NiN.wetland.cov) %>% 
  kable("html", caption = "Reference data set") %>% 
  kable_styling("striped") %>% scroll_box(width = "100%")
```

@@

```{r, attr.output='style="max-height: 300px;"', results='hide', class.source = 'fold-hide'}

@@

We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021).
```{r, results='hide', class.source = 'fold-hide'}

@@

```{r, eval = F, class.source = 'fold-hide'}

@@

### Ecosystem sub-types {#ecosystem-sub-types-fpiw}
And we can show the resulting scaled values as Violin plots for each indicator and main ecosystem type 
```{r fpiw-plot-maintype, message=FALSE, warning=FALSE, fig.cap="Scaled vegetation indicator values for wetlands. 1 = cases where variable is lower than the reference value- 2 = cases where variablei is higher than the reference value.", out.width='100%', class.source = 'fold-hide'}

# making the plot
ggplot(res.wetland, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
  geom_hline(yintercept=0.6, linetype="dashed") + 
  geom_violin() +
#  geom_boxplot(width=0.2, color="grey") +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
  xlab("Wetland ecosystem sub-type") + 
  ylab("Scaled indicator value") 
```

This figure is a bit easier to read if we remove the sub-types with very few data points:
```{r fpiw-plot-maintype-v2, message=FALSE, warning=FALSE, fig.cap="Scaled vegetation indicator values for the four most common sub-types of wetland. 1 = cases where variable is lower than the reference value- 2 = cases where variablei is higher than the reference value.", out.width='100%', class.source = 'fold-hide'}
ggplot(res.wetland[res.wetland$hovedtype_rute %in% c("V1", "V2", "V3", "V4"),], aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
  geom_hline(yintercept=0.6, linetype="dashed") + 
  geom_violin() +
#  geom_boxplot(width=0.2, color="grey") +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
  xlab("Wetland ecosystem sub-type") + 
  ylab("Scaled indicator value") +
  guides(fill='none')
```

@@

```{r ph-lower-map-fpiw, warning=FALSE, message = FALSE, fig.cap="Functional plant indicator values for pH (lower) in wetlands.", class.source = 'fold-hide'}
#### scaled value maps ####
# keep wide format and add geometry again
res.wetland2 <- results.wetland[['2-sided']]
st_geometry(res.wetland2) <- st_geometry(ANO.wetland)

@@

### Unscaled values vs. reference
We can also compare the unscaled values to the reference distribution in order to identify ecosystem types and functional plant indicators showing a deviation from the expectation. Since pH and nitrogen show some deviation in the lower indicator we exemplify this with these two indicators for unscaled values.

```{r pH-distcomp-fpiw, class.source = 'fold-hide', fig.cap="Comparing the reference and variable distribution for the pH indicator in ecosystem sub-type V1."}
    #summary(res.wetland$kartleggingsenhet_1m2)
    #length(unique(res.wetland$kartleggingsenhet_1m2))
# 16 NiN-types to plot
    #colnames(wetland.ref.cov[['Soil_reaction_pH']])

@@

```{r ph-distcomp2-fpiw, class.source = 'fold-hide', fig.cap="Comparing the reference and variable distribution for the pH indicator in ecosystem sub-types other than V1."}
par(mfrow=c(2,4))
## V2s
# V2-C-1
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C1a","V2-C1b")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V2-C1',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original)),
       col="red")

@@

```{r n-distcomp-fpiw, class.source = 'fold-hide', fig.cap="Comparing the reference and variable distribution for the nitrogen affiliation indicator in ecosystem sub-type V1."}
### Nitrogen
par(mfrow=c(2,4))
## V1s
# V1-C-1
plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C1a","V1-C1b","V1-C1c","V1-C1d","V1-C1e")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V1-C1',xlab='Nitrogen value')
points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original)),
       col="red")

@@

```{r n-distcomp2-fpiw,class.source = 'fold-hide', fig.cap="Comparing the reference and variable distribution for the nitrogen affiliation indicator in ecosystem sub-types other than V1."}
par(mfrow=c(2,4))
## V2s
# V2-C-1
plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V2-C1a","V2-C1b")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V2-C1',xlab='Nitrogen value')
points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original)),
       col="red")

>>>>

@@

<!-- Load all you dependencies here -->
```{r setup, include=FALSE, warning=F, message=F}
library(sf)
library(plyr)
library(tidyverse)
library(RColorBrewer)
library(gridExtra) 
library(ggridges)
library(tmap)
library(tmaptools)
library(raster)
library(stars)
library(knitr)
library(kableExtra)
library(betareg)
library(glmmTMB)
knitr::opts_chunk$set(echo = TRUE)
```

@@

Indicators described in this chapter:

@@

## Analyses {#analyses-fpiw}
### Data sets
<!-- Please print the entire workflow with echo=T. Also, simply reading cached data creates loose ends - I cant find out where this data comes from. You may also chose to read tin this data (or a cached version of it) way at the top of the file, so that you can add some in-line R code for the summary stats in the introduction part (see previous comments)-->

Import ANO data:
```{r}
# Add data from cache
ANO.sp<-readRDS(paste0(here::here(),"/data/cache/ANO.sp.RDS"))
ANO.geo<-readRDS(paste0(here::here(),"/data/cache/ANO.geo.RDS"))

@@

Import Plant indicators from Tyler et al. (2021):
```{r}
## Tyler indicator data
## data from cache
ind.dat <-readRDS(paste0(here::here(),"/data/cache/ind.dat.RDS"))
# ind.dat <- read.table("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/functional plant indicators/Tyler et al_Swedish plant indicators.txt",
#                         sep = '\t', header=T, quote = '')
# head(ind.dat)

@@

Import Generalized species lists NiN
```{r}
# data from cache
Eco_State<-readRDS(paste0(here::here(),"/data/cache/Eco_State.RDS"))
# load("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/functional plant indicators/reference from NiN/Eco_State.RData")
# str(Eco_State)
```

@@

```{r, include = T, results='hide', warning=F, message=F}

@@

<!-- Print head of data set with horizontal scrolling -->
```{r ano-sp-tab, echo=F}
head(ANO.sp.ind) %>%
  kable("html",caption = "ANO species occurence data set with attached plant trait data.") %>% 
  kable_styling("striped") %>% scroll_box(width = "100%")
```

```{r wetland-ref-data-tab, echo=FALSE}
head(NiN.wetland.cov) %>% 
  kable("html", caption = "Reference data set") %>% 
  kable_styling("striped") %>% scroll_box(width = "100%")
```

@@

```{r, attr.output='style="max-height: 300px;"', results='hide'}

@@

We then can prepare a list of data frames to hold the results and perform the scaling according to the principles described in NINA report 1967 (Töpper and Jakobsson 2021).
```{r, results='hide'}

@@

```{r, eval = F}

@@

### Ecosystem sub-types {#ecosystem-sub-types-fpiw}
And we can show the resulting scaled values as Violin plots for each indicator and main ecosystem type 
```{r fpiw-plot-maintype, message=FALSE, warning=FALSE, fig.cap="Scaled vegetation indicator values for wetlands. 1 = cases where variable is lower than the reference value- 2 = cases where variablei is higher than the reference value.", out.width='100%', echo=F}
# making the plot
ggplot(res.wetland, aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
  geom_hline(yintercept=0.6, linetype="dashed") + 
  geom_violin() +
#  geom_boxplot(width=0.2, color="grey") +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
  xlab("Wetland ecosystem sub-type") + 
  ylab("Scaled indicator value") 
```

This figure is a bit easier to read if we remove the sub-types with very few data points:
```{r fpiw-plot-maintype-v2, message=FALSE, warning=FALSE, fig.cap="Scaled vegetation indicator values for the four most common sub-types of wetland. 1 = cases where variable is lower than the reference value- 2 = cases where variablei is higher than the reference value.", out.width='100%'}
ggplot(res.wetland[res.wetland$hovedtype_rute %in% c("V1", "V2", "V3", "V4"),], aes(x=factor(hovedtype_rute), y=scaled_value, fill=fp_ind)) + 
  geom_hline(yintercept=0.6, linetype="dashed") + 
  geom_violin() +
#  geom_boxplot(width=0.2, color="grey") +
  geom_point(size=0.7, shape=16, color="grey") +
  facet_wrap(~factor(fp_ind,levels=c("Light1","Moist1","pH1","Nitrogen1","Light2","Moist2","pH2","Nitrogen2")), ncol = 4) + 
  xlab("Wetland ecosystem sub-type") + 
  ylab("Scaled indicator value") +
  guides(fill='none')
```

@@

```{r ph-lower-map-fpiw, warning=FALSE, message = FALSE, fig.cap="Functional plant indicator values for pH (lower) in wetlands."}
#### scaled value maps ####
# keep wide format and add geometry again
res.wetland2 <- results.wetland[['2-sided']]
st_geometry(res.wetland2) <- st_geometry(ANO.wetland)

@@

### Unscaled values vs. reference
We can also compare the unscaled values to the reference distribution in order to identify ecosystem types and functional plant indicators showing a deviation from the expectation. Since pH and nitrogen show some deviation in the lower indicator we exemplify this with these two indicators for unscaled values. (Only the code for the first figure is shown).

```{r pH-distcomp-fpiw, fig.cap="Comparing the reference and variable distribution for the pH indicator in ecosystem sub-type V1."}
    #summary(res.wetland$kartleggingsenhet_1m2)
    #length(unique(res.wetland$kartleggingsenhet_1m2))
# 16 NiN-types to plot
    #colnames(wetland.ref.cov[['Soil_reaction_pH']])

@@

```{r ph-distcomp2-fpiw, fig.cap="Comparing the reference and variable distribution for the pH indicator in ecosystem sub-types other than V1.", echo=FALSE}
par(mfrow=c(2,4))
## V2s
# V2-C-1
plot(density( as.matrix(wetland.ref.cov[['Soil_reaction_pH']][,c("V2-C1a","V2-C1b")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V2-C1',xlab='pH value')
points(density(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="pH1" & res.wetland$fp_ind=="pH1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original)),
       col="red")

@@

```{r n-distcomp-fpiw, fig.cap="Comparing the reference and variable distribution for the nitrogen affiliation indicator in ecosystem sub-type V1.", echo=F}
### Nitrogen
par(mfrow=c(2,4))
## V1s
# V1-C-1
plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V1-C1a","V1-C1b","V1-C1c","V1-C1d","V1-C1e")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V1-C1',xlab='Nitrogen value')
points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V1-C-1",]$original)),
       col="red")

@@

```{r n-distcomp2-fpiw, fig.cap="Comparing the reference and variable distribution for the nitrogen affiliation indicator in ecosystem sub-types other than V1.", echo=F}
par(mfrow=c(2,4))
## V2s
# V2-C-1
plot(density( as.matrix(wetland.ref.cov[['Nitrogen']][,c("V2-C1a","V2-C1b")]) ,na.rm=T),
     xlim=c(1,7), type="l", main='V2-C1',xlab='Nitrogen value')
points(density(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,na.rm=T),
       type="l", col="red")
points(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original,
       rep(0,length(res.wetland[res.wetland$fp_ind=="Nitrogen1" & res.wetland$fp_ind=="Nitrogen1" & res.wetland$kartleggingsenhet_1m2=="V2-C-1",]$original)),
       col="red")
